---
title: "Map COVID PBMC datasets to a healthy reference"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---


```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = 'styler',
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  time_it = TRUE,
  error = TRUE
)
```

```{r, warning=F, message=F}
library(Seurat)
library(BPCells)
library(dplyr)
library(patchwork)
library(ggplot2)
options(future.globals.maxSize = 1e9)
```


## Introduction: Reference mapping analysis in Seurat v5

In Seurat v5, we introduce a scalable approach for reference mapping datasets from separate studies or individuals. Reference mapping is a powerful approach to identify consistent labels across studies and perform cross-dataset analysis. We emphasize that while individual datasets are manageable in size, the aggregate of many datasets often amounts to millions of cell which do not fit in-memory. Furthermore, cross-dataset analysis is often challenged by disparate or unique cell type labels. Through reference mapping, we annotate all cells with a common reference for consistent cell type labels. Importantly, we never simultaneously load all of the cells in-memory to maintain low memory usage.
 
In this vignette, we reference map three publicly available datasets totaling 1,498,064 cells and 277 donors which are available through [CZI cellxgene collections](https://cellxgene.cziscience.com/collections): [Ahern, et al., Nature 2022](https://cellxgene.cziscience.com/collections/8f126edf-5405-4731-8374-b5ce11f53e82), [Jin, et al., Science 2021](https://cellxgene.cziscience.com/collections/b9fc3d70-5a72-4479-a046-c2cc1ab19efc), and [Yoshida, et al., Nature 2022](https://cellxgene.cziscience.com/collections/03f821b4-87be-4ff4-b65a-b5fc00061da7). Each dataset consists of PBMCs from both healthy donors and donors diagnosed with COVID-19. Using the harmonized annotations, we demonstrate how to prepare a pseudobulk object to perform differential expression analysis across disease within cell types.

Prior to running this vignette, please [install Seurat v5](install.html) and see the [BPCells vignette](seurat5_bpcells_interaction_vignette.html) to construct the on-disk object used in this vignette. Additionally, we map to our annotated CITE-seq reference containing 162,000 cells and 228 antibodies ([Hao*, Hao*, et al., Cell 2021](https://doi.org/10.1016/j.cell.2021.04.048)) which is available for download [here](https://zenodo.org/record/7779017#.ZCMojezMJqs).

## Load the PBMC Reference Dataset and Query Datasets
We first load the reference (available [here](https://zenodo.org/record/7779017#.ZCMojezMJqs)) and normalize the query Seurat object prepared in the [BPCells interaction vignette](seurat5_bpcells_interaction_vignette.html). The query object consists of datasets from three different studies constructed using the `CreateSeuratObject` function, which accepts a list of BPCells matrices as input. Within the Seurat object, the three datasets reside in the `RNA` assay in three separate `layers` on-disk.


```{r load.data}
reference <- readRDS("/brahms/hartmana/vignette_data/pbmc_multimodal_2023.rds")
object <- readRDS("/brahms/hartmana/vignette_data/merged_covid_object.rds")
object <- NormalizeData(object, verbose = FALSE)
```

## Mapping
Using the same code from the [v4 reference mapping vignette](articles/multimodal_reference_mapping.html), we find anchors between the reference and query in the precomputed supervised PCA. We recommend the use of supervised PCA for CITE-seq reference datasets, and demonstrate how to compute this transformation in [v4 reference mapping vignette](articles/multimodal_reference_mapping.html). In Seurat v5, we only need to call `FindTransferAnchors` and `MapQuery` once to map all three datasets as they are all contained within the query object. Furthermore, utilizing the on-disk capabilities of [BPCells](https://github.com/bnprks/BPCells), we map 1.5 million cells without ever loading them all into memory.  

```{r}
anchor <- FindTransferAnchors(
  reference = reference,
  query = object,
  reference.reduction = 'spca',
  normalization.method = 'SCT',
  dims = 1:50)
object <- MapQuery(
  anchorset = anchor,
  query = object,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2"
    ),
  reduction.model = 'wnn.umap'
)
```

## Explore the mapping results
Next, we visualize all cells from the three studies which have been projected into a UMAP-space defined by the reference. Each cell is annotated at two levels of granularity (`predicted.celltype.l1` and `predicted.celltype.l2`). We can compare the differing ontologies used in the original annotations (`cell_type`) to the now harmonized annotations (`predicted.celltype.l2`, for example) that were predicted from reference-mapping. Previously, the lack of standardization prevented us from directly performing integrative analysis across studies, but now we can easily compare.

```{r, fig.width=10, fig.height=6}
DimPlot(object, reduction = 'ref.umap', group.by = 'cell_type',alpha = 0.1, label = TRUE, split.by = 'publication', ncol = 3,  label.size = 3) + NoLegend()
```
```{r, fig.width=10, fig.height=6}
DimPlot(object, reduction = 'ref.umap', group.by = 'predicted.celltype.l2',alpha = 0.1, label = TRUE, split.by = 'publication', ncol = 3,  label.size = 3) + NoLegend()
```

## Differential composition analysis
We utilize our harmonized annotations to identify differences in the proportion of different cell types between healthy individuals and COVID-19 patients. For example, we noticed a reduction in MAIT cells as well as an increase in plasmablasts among COVID-19 patients.

```{r}
df_comp <- as.data.frame.matrix(table(object$donor_id, object$predicted.celltype.l2))
select.donors <- rownames(df_comp)[rowSums(df_comp)> 50]
df_comp <- df_comp[select.donors, ]
df_comp_relative <- sweep(x = df_comp, MARGIN = 1, STATS = rowSums(df_comp), FUN = '/')

df_disease <-  as.data.frame.matrix(table(object$donor_id, object$disease))[select.donors, ]

df_comp_relative$disease <- 'other'
df_comp_relative$disease[df_disease$normal!=0] <- 'normal'
df_comp_relative$disease[df_disease$`COVID-19`!=0] <- 'COVID-19'
df_comp_relative$disease <- factor(df_comp_relative$disease, levels = c('normal','COVID-19','other'))
df_comp_relative <- df_comp_relative[df_comp_relative$disease %in% c('normal','COVID-19'),]
```

```{r, fig.width=10, fig.height=4}
p1 <-  ggplot(data = df_comp_relative, mapping = aes(x = disease, y = MAIT, fill = disease)) +  
  geom_boxplot(outlier.shape  = NA) +
  scale_fill_manual(values = c("#377eb8", "#e41a1c")) +
  xlab("") + ylab('relative abundance') +
  ggtitle('MAIT') +
  geom_jitter(color="black", size=0.4, alpha=0.9 ) +
  theme_bw() +
  theme( axis.title = element_text(size = 12),
         axis.text = element_text(size = 12),
         plot.title = element_text(size = 15, hjust = 0.5, face = "bold")
  )

p2 <-  ggplot(data = df_comp_relative, mapping = aes(x = disease, y = Plasmablast, fill = disease)) +  
  geom_boxplot(outlier.shape  = NA) +
  scale_fill_manual(values = c("#377eb8", "#e41a1c")) +
  xlab("") + ylab('relative abundance') +
    ggtitle('Plasmablast') +
  geom_jitter(color="black", size=0.4, alpha=0.9 ) +
  theme_bw() +
  theme( axis.title = element_text(size = 12),
         axis.text = element_text(size = 12),
         plot.title = element_text(size = 15, hjust = 0.5, face = "bold")
  )

p1 + p2 + plot_layout(ncol = 2)
```

## Differential expression analysis
In addition to composition analysis, we use an aggregation-based (pseudobulk) workflow to explore differential genes between healthy individuals and COVID-19 donors. We aggregate all cells within the same cell type and donor using the `AggregateExpression` function. This returns a Seurat object where each ‘cell’ represents the pseudobulk profile of one cell type in one individual.

```{r}
bulk <- AverageExpression(object,
  method = 'aggregate',
  return.seurat = TRUE,
  slot = 'counts',
  assays  = 'RNA',
  group.by = c("predicted.celltype.l2", "donor_id", "disease")
)

bulk$celltype <- sapply(strsplit(Cells(bulk), split = "_"), '[', 1)
bulk$donor <- sapply(strsplit(Cells(bulk), split = "_"), '[', 2)
bulk$disease <- sapply(strsplit(Cells(bulk), split = "_"), '[', 3)
```

```{r} 
bulk <- subset(bulk, subset = disease %in% c('normal', 'COVID-19') )
bulk <- subset(bulk, subset = celltype !=  c('Doublet') )
bulk$disease <- factor(bulk$disease, levels = c('normal', 'COVID-19'))
```

Once a pseudobulk object is created, we can perform cell type-specific differential expression analysis between healthy individuals and COVID-19 donors. Here, we only visualize certain interferon-stimulated genes which are often upregulated during viral infection.

```{r, fig.width=10, fig.height=12}
p1 <- VlnPlot(
  object = bulk, features = 'IFI6', group.by = 'celltype',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p2 <- VlnPlot(
  object = bulk, features = c('ISG15'), group.by = 'celltype',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p3 <- VlnPlot(
  object = bulk, features = c('IFIT5'), group.by = 'celltype',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p1 + p2 + p3 + plot_layout(ncol = 1)
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
