---
title: "Seurat Command List"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 120),
  message = FALSE,
  warning = FALSE,
  results = 'hold',
  eval = FALSE,
  error = TRUE
)
```

```{r load-data, echo=FALSE}
library(Seurat)
library(ggplot2)
library(SeuratData)

# install datasets 
InstallData('pbmc3k')
InstallData('cbmc')
InstallData('ifnb')

#load in pbmc example
pbmc <- LoadData('pbmc3k')

# load in multimodal example
# type ?cbmc for more info
cbmc <- LoadData('cbmc')

#load in case/control example
# type ?ifnb for more info
ifnb <- LoadData('ifnb')

```

# Standard `Seurat` workflow
```{r seurat-standard-workflow-2}
pbmc <- NormalizeData(object = pbmc)
pbmc <- FindVariableFeatures(object = pbmc)
pbmc <- ScaleData(object = pbmc)
pbmc <- RunPCA(object = pbmc)
pbmc <- FindNeighbors(object = pbmc, dims = 1:30)
pbmc <- FindClusters(object = pbmc)
pbmc <- RunUMAP(object = pbmc, dims = 1:30)
DimPlot(object = pbmc, reduction = 'umap')

```

## SCtransform version
```{r seurat-standard-workflow}
pbmc <- SCTransform(object = pbmc)
pbmc <- RunPCA(object = pbmc)
pbmc <- FindNeighbors(object = pbmc, dims = 1:30)
pbmc <- FindClusters(object = pbmc)
pbmc <- RunUMAP(object = pbmc, dims = 1:30)
DimPlot(object = pbmc, reduction = 'umap')
```

```{r chaincmds,eval=FALSE}
# note that you can chain multiple commands together with %>%
pbmc <- SCTransform(pbmc) %>% RunPCA() %>% FindNeighbors(dims=1:30) %>% 
  FindClusters() %>% RunUMAP(dims=1:30)
```

# `Seurat` Object Data Access

## Cell, feature, and layer names
```{r features-and-cells}
# Get cell and feature names, and total numbers
# We show multiple ways to get the same output
# cell names
colnames(pbmc)
Cells(pbmc)

# feature names
Features(pbmc)
rownames(pbmc)

# number of cells/features
num_cells <- ncol(pbmc)
num_features <- nrow(pbmc)

# List of object  layers
Layers(pbmc)

# working with multimodal objects
# list assays
Assays(cbmc)

# Assay-specific features (genes/ADT)
Features(cbmc[["RNA"]])
Features(cbmc[["ADT"]])

# Variable feature names
VariableFeatures(pbmc)

```

```{r setvarfeatures, eval=FALSE}
# Set variable features
VariableFeatures(cbmc) <- var.gene.names

# set for a specific assay
VariableFeatures(cbmc[["ADT"]]) <- var.gene.names
```

## Identity class labels
```{r idents}
# Setting and retrieving cell identities

# Set identity classes to an existing column in meta data
Idents(object = pbmc) <- 'seurat_annotations'

# View cell identities, get summary table 
Idents(pbmc)
table(Idents(pbmc))

# Set identity to CD4 T cells for all cells
Idents(pbmc) <- 'CD4 T cells'

# Set for a selected group of cells
pbmc.cells <- Cells(pbmc)
Idents(object = pbmc, cells = pbmc.cells[1:10]) <- 'CD4 T cells'

# Get cell identity classes
Idents(object = pbmc)
levels(x = pbmc)

# Stash cell identity classes in metadata
pbmc[['old.ident']] <- Idents(object = pbmc)
pbmc <- StashIdent(object = pbmc, save.name = 'old.ident')

# Rename identity classes
pbmc <- RenameIdents(object = pbmc, 'CD4 T cells' = 'T Helper cells')
```

## Cell metadata
```{r metadata}
# View metadata data frame, stored in object@meta.data
pbmc[[]]

# Retrieve specific values from the metadata
pbmc$nCount_RNA
pbmc[[c('percent.mito', 'nFeature_RNA')]]

# Add metadata, see ?AddMetaData
random_group_labels <- sample(x = c('g1', 'g2'), size = ncol(x = pbmc), replace = TRUE)
pbmc$groups <- random_group_labels
```

## Expression data (stored as layers in Seurat v5)

```{r expression-matrices, eval=FALSE}
# Retrieve data in an expression matrix 
# RNA counts matrix 
pbmc[["RNA"]]$counts

# Alternate accessor function with the same result
LayerData(pbmc,assay = 'RNA',layer = 'counts')

# GetAssayData from Seurat v4 is still supported
GetAssayData(object = pbmc, assay = 'RNA', layer = 'counts')

# ADT counts matrix (multimodal object)
cbmc[["ADT"]]$counts

```

```{r setdata, eval = FALSE}
# Set expression data  
# assume new.data is a new expression matrix
pbmc[["RNA"]]$counts <- new.data

# Alternate setter function with the same result
LayerData(pbmc,assay = 'RNA',layer = 'counts') <- new.data

# SetAssayData from Seurat v4 is still supported
pbmc <- SetAssayData(object = pbmc, layer = 'counts',new.data = new.data)

```

## Dimensional reductions
```{r embeddings-loadings}
# Get cell embeddings and feature loadings
# stored on pbmc[["pca"]]@cell.embeddings
Embeddings(pbmc, reduction = 'pca')

# stored in pbmc[["pca]]@feature.loadings
Loadings(pbmc, reduction = 'pca')
```

```{r customreduc, eval=FALSE}
# Create custom dimensional reduction
# loadings matrix is optional
new_reduction <- CreateDimReducObject(embeddings = new.embeddings,loadings = new.loadings, key = "custom_pca")
pbmc[["custom_pca"]] <- new_reduction
```

## FetchData 
```{r fetchdata}
# FetchData can access anything from expression matrices, cell embeddings, or metadata
# Use the previously listed commands to access entire matrices
# Use FetchData to access individual/small groups of variables
FetchData(object = pbmc, vars = c('PC_1', 'nFeature_RNA', 'MS4A1'),layer = 'counts')
```

# Subsetting and merging

## Subset Seurat objects
```{r subsetting}
# Subset Seurat object based on identity class, also see ?SubsetData
subset(x = pbmc, idents = 'B')
subset(x = pbmc, idents = c('Naive CD4 T', 'CD8 T'), invert = TRUE)

# Subset on the expression level of a gene/feature
subset(x = pbmc, subset = MS4A1 > 2.5)

# Subset on a combination of criteria
subset(x = pbmc, subset = MS4A1 > 2.5 & PC_1 > 5)
subset(x = pbmc, subset = MS4A1 > 2.5, idents = 'B')

# Subset on a value in the object meta data
subset(x = pbmc, subset = groups == "g1")

# Downsample the number of cells per identity class
subset(x = pbmc, downsample = 100)

```

## Split layers
```{r splitlayers, results = 'hide'}
# In Seurat v5, users can now split in object directly into different layers
# keeps expression data in one object, but splits multiple samples into layers
# can proceed directly to integration workflow after splitting layers
ifnb[["RNA"]] <- split(ifnb[["RNA"]],f = ifnb$stim)
Layers(ifnb)

# If desired, for example after intergation, the layers can be joined together again
ifnb <- JoinLayers(ifnb)
```

## Split objects
```{r splitting, results = 'hide'}
# In line with prior workflows, you can also into split your object into a list of multiple objects
# based on a metadata column 
# creates a list of two objects 
ifnb_list <- SplitObject(ifnb,split.by = 'stim')
ifnb_list$CTRL
ifnb_list$STIM
```

## Merge objects (without integration)
In Seurat v5, merging creates a single object, but keeps the expression information split into different layers for integration. If not proceeding with integration, rejoin the layers after merging.

```{r merging-2, eval=FALSE, results = 'hide'}
# Merge two Seurat objects
merged_obj <- merge(x = ifnb_list$CTRL, y = ifnb_list$STIM)
merged_obj[["RNA"]] <- JoinLayers(merged_obj)

# Example to merge more than two Seurat objects
merge(x = pbmc1, y = list(pbmc2, pbmc3))
```

## Merge objects (with integration)
See [introduction to integration](integration_introduction.html) for more information.

```{r merging-3, eval=FALSE, results = 'hide'}
merged_obj <- merge(x = ifnb_list$CTRL, y = ifnb_list$STIM)
merged_obj <- NormalizeData(merged_obj)
merged_obj <- FindVariableFeatures(merged_obj)
merged_obj <- ScaleData(merged_obj)
merged_obj <- RunPCA(merged_obj)
merged_obj <- IntegrateLayers(
  object = obj, method = RPCAIntegration,
  orig.reduction = "pca", new.reduction = 'integrated.rpca',
  verbose = FALSE)

# now that integration is complete, rejoin layers
merged_obj[["RNA"]] <- JoinLayers(merged_obj)

```

# Pseudobulk analysis

## Group cells together, based on multiple categories
See [DE vignette](de_vignette.html) for information  on how to add the `donor_id` column to meta data.

```{r, eval=FALSE}
# pseudobulk cells only by cell type
bulk <- AggregateExpression(ifnb,group.by ='seurat_annotations', return.seurat = TRUE)
Cells(bulk)

# pseudobulk cells by stimulation condition AND cell type
bulk <- AggregateExpression(ifnb,group.by = c('stim','seurat_annotations'), return.seurat = TRUE)
Cells(bulk)

# pseudobulk cells by stimulation condition AND cell type AND donor
bulk <- AggregateExpression(ifnb,group.by = c('stim','seurat_annotations',"donor_id"), return.seurat = TRUE)
Cells(bulk)
```

# Visualization in `Seurat` 

Seurat has a vast, ggplot2-based plotting library. All plotting functions will return a ggplot2 plot by default, allowing easy customization with ggplot2.

```{r visualization}

# Dimensional reduction plot 
DimPlot(object = pbmc, reduction = 'pca')

# Dimensional reduction plot, with cells colored by a quantitative feature
# Defaults to UMAP if available
FeaturePlot(object = pbmc, features = "MS4A1")

# Scatter plot across single cells
FeatureScatter(object = pbmc, feature1 = "MS4A1", feature2 = "PC_1")
FeatureScatter(object = pbmc, feature1 = "MS4A1", feature2 = "CD3D")

# Scatter plot across individual features, repleaces CellPlot
CellScatter(object = pbmc, cell1 = "AGTCTACTAGGGTG", cell2 = "CACAGATGGTTTCT")

VariableFeaturePlot(object = pbmc)

#Violin and Ridge plots
VlnPlot(object = pbmc, features = c("LYZ", "CCL5", "IL32"))
RidgePlot(object = pbmc, feature = c("LYZ", "CCL5", "IL32"))
```

```{r heatmaps}
# Heatmaps (visualize scale.data slot)
DimHeatmap(object = pbmc,reduction = 'pca', cells = 200)

# standard workflow
pbmc <- ScaleData(pbmc,features = heatmap_markers)
DoHeatmap(object = pbmc,features = heatmap_markers)

# sctransform workflow
pbmc <- GetResidual(pbmc,features = heatmap_markers)
DoHeatmap(object = pbmc,features = heatmap_markers)

# heatmap with maximum of 100 cells per group
DoHeatmap(pbmc,heatmap_markers,cells = subset(pbmc,downsample = 100))

```

```{r newthings}
# New things to try! 
# Note that plotting functions now return ggplot2 objects, so you can add themes, titles, and options onto them
VlnPlot(object = pbmc, features = "MS4A1", split.by = "groups")
DotPlot(object = pbmc, features = c("LYZ", "CCL5", "IL32"), split.by = "groups")
FeaturePlot(object = pbmc, features = c("MS4A1", "CD79A"), blend = TRUE)
DimPlot(object = pbmc) + DarkTheme()
DimPlot(object = pbmc) + labs(title = '2,700 PBMCs clustered using Seurat and viewed\non a two-dimensional UMAP')
```

Seurat provides many prebuilt themes that can be added to ggplot2 plots for quick customization

| Theme | Function |
| ----- | -------- |
| `DarkTheme` | Set a black background with white text |
| `FontSize` | Set font sizes for various elements of a plot |
| `NoAxes` | Remove axes and axis text |
| `NoLegend` | Remove all legend elements |
| `RestoreLegend` | Restores a legend after removal |
| `RotatedAxis` | Rotates x-axis labels |

```{r helper-functions}

# Plotting helper functions work with ggplot2-based scatter plots, such as DimPlot, FeaturePlot, CellScatter, and FeatureScatter
plot <- DimPlot(object = pbmc) + NoLegend()

# HoverLocator replaces the former `do.hover` argument
# It can also show extra data throught the `information` argument, designed to work smoothly with FetchData
HoverLocator(plot = plot, information = FetchData(object = pbmc, vars = c("ident", "PC_1", "nFeature_RNA")))

# FeatureLocator replaces the former `do.identify`
select.cells <- FeatureLocator(plot = plot)

# Label points on a ggplot object
LabelPoints(plot = plot, points = TopCells(object = pbmc[["pca"]]), repel = TRUE)
```

# Multi-Assay Features

With Seurat, you can easily switch between different assays at the single cell level (such as ADT counts from CITE-seq, or integrated/batch-corrected data). Most functions now take an assay parameter, but you can set a Default Assay to avoid repetitive statements.

```{r multi-assay}
cbmc <- CreateSeuratObject(counts = cbmc.rna)
# Add ADT data
cbmc[['ADT']] <- CreateAssayObject(counts = cbmc.adt)
# Run analyses by specifying the assay to use
NormalizeData(object = cbmc, assay = 'RNA')
NormalizeData(object = cbmc, assay = 'ADT', method = 'CLR')

# Retrieve and set the default assay
DefaultAssay(object = cbmc)
DefaultAssay(object = cbmc) <- 'ADT'
DefaultAssay(object = cbmc)

# Pull feature expression from both assays by using keys
FetchData(object = cbmc, vars = c('rna_CD3E', 'adt_CD3'))

# Plot data from multiple assays using keys
FeatureScatter(object = cbmc, feature1 = "rna_CD3E", feature2 = "adt_CD3")
```

# Additional resources

Users who are particularly interested in some of the technical changes to data storage in Seurat v5 can explore the following resources:

* [SeuratObject manual](https://mojaveazure.github.io/seurat-object/)
* [Seurat v5 and Assay5 introductory vignette ](seurat5_essential_commands.html)

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
