---
title: "Getting started with Seurat 5"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
---

```{r fxns, include = FALSE}
library('htmlTable')
make_list <- function(items) {
  paste0("<ul>", sprintf('<li>%s</li>', items), '</ul>', collapse = '')
}
make_href <- function(url, text){
  paste0("<div style='text-align:center; font-size:large; font-weight:bold'><a href=", url ,">", text, "</a></div>")
}
make_href2 <- function(url, text){
  paste0("<a href=", url ,">", text, "</a>")
}
process_entry <- function(dat) {
  if (grepl(pattern = "https://satijalab.org/img/vignette_images", x = dat$image)) {
    img <- paste0('![](', dat$image, '){width=3000px}')
  } else if (grepl(pattern = "assets/", x= dat$image)) {
    img <- paste0('![](', dat$image, '){width=3000px}') 
  } else {
    img <- paste0('![](', '../output/images/', dat$image, '){width=3000px}')
  }

  if (dat$name == "seurat5_run_azimuth") {
    link <- "https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html"
  } else if (grepl(pattern = "https://satijalab.org/", x = dat$name)) {
    link <- dat$name
  } else {
    link <- paste0(dat$name, ".html")
  }
  go.button <- paste0('<a class="btn btn-primary" href="', link, '" role="button" style="width:100%">GO</a>')
  data.frame(
    title = make_href(url = link, text = dat$title),
    img = img,
    desc = dat$summary,
    btn = go.button
  )
}
process_wrapper_entry <- function(dat) {
  data.frame(
    Package = dat$name,
    Vignette = make_href2(url = dat$link, text = dat$title),
    Reference = make_href2(url = dat$reference, text = dat$citation),
    Source = make_href2(url = dat$source, text = dat$source)
  )
}
make_vignette_card_section <- function(vdat, cat) {
  vignettes <- vdat[[cat]]$vignettes
  dat <- data.frame(title = character(), img = character(), desc = character())
  for (v in 1:length(x = vignettes)) {
    dat <- rbind(dat, process_entry(vignettes[[v]]))
    if(nrow(x = dat) == 3 | v == length(x = vignettes)){
      colnames(dat) <- NULL
      dat <- t(dat)
      if (ncol(x = dat) == 2) {
        print(htmlTable(
          dat, 
          align = '|l|l|', 
          css.cell = "padding-left: .75em; width: 50%", 
          css.class = "two-column-htmltable"
        ))
      } else if (ncol(x = dat) == 1){
        print(htmlTable(
          dat, 
          align = '|l|', 
          css.cell = "padding-left: .75em; width: 100%", 
          css.class = "one-column-htmltable"
        ))
      } else {
        print(htmlTable(
          dat, 
          align = '|l|l|l|', 
          css.cell = "padding-left: .75em; width: 30%"
        ))
      }
      dat <- data.frame(title = character(), img = character(), desc = character())
    }
  }
}
```

```{r yaml, include = FALSE}
library(yaml)
vdat <- read_yaml(file = "vignettes_v5.yaml")
```

```{=html}
<style>
.two-column-htmltable {
  width: 66.666%;
}
.one-column-htmltable {
  width: 33.333%;
}

</style>
```

# Introduction to Seurat 5

The following vignettes outline how to install, create, and interact with Seurat 5 objects.

```{r results='asis', echo=FALSE, warning=FALSE, message = FALSE}
make_vignette_card_section(vdat = vdat, cat = 1)
```

# Analysis in Seurat 5

Below we demonstrate new functionality introduced in Seurat 5 including:

* Sketch-based workflows to efficiently process large datasets while preserving rare and abundant cell types
* Bridge integration to annotate cells from complementary technologies with a scRNA-seq reference
* Spatial infrastructure to analyze public datasets from multiple technologies
* Integration using CCA, RPCA, scVI, and Harmony in a common framework

```{r results='asis', echo=FALSE, warning=FALSE, message = FALSE}
make_vignette_card_section(vdat = vdat, cat = 2)
```
