---
title: "Installation Instructions for Seurat"
output: html_document
---

To install Seurat, [R](https://www.r-project.org/) version 4.0 or greater is required. We also recommend installing [R Studio](https://www.rstudio.com/).

# Install from CRAN 
 
Seurat is available on [CRAN](https://cran.r-project.org/package=Seurat) for all platforms. To install, run: 

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
install.packages('Seurat')
library(Seurat)
```

If you see the warning message below, enter `y`:

```{r eval=FALSE}
package which is only available in source form, and may need compilation of C/C++/Fortran: 'Seurat'
Do you want to attempt to install these from sources?
y/n:
```

# Install previous versions of Seurat

## Install any version 3 release

Any of the Seurat version 3 releases can be installed with the following command:

```{r eval = FALSE}
remotes::install_version("Seurat", version = "3.X.X")
```

## Install the last version 2 release (2.3.4) 

To facilitate easy re-installation of the last version 2 release, we are hosting the binaries on our website. These can be installed with the following command: 

```{r eval = FALSE}
source("https://z.umn.edu/archived-seurat")
```

 <details>
  <summary>View the script</summary>
  <script src="https://gist.github.com/satijalab/beb9bb50dedc75ee023bd5d9be5fe684.js"></script>
</details>

## Older versions of Seurat 
Old versions of Seurat, from Seurat v2.0.1 and up, are hosted in CRAN's archive. To install an old version of Seurat, run: 

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
# Install the remotes package 
install.packages('remotes')
# Replace '2.3.0' with your desired version
remotes::install_version(package = 'Seurat', version = package_version('2.3.0'))
library(Seurat)
```

For versions of Seurat older than those not hosted on CRAN (versions 1.3.0 and 1.4.0), please download the packaged source code from our [releases page](https://github.com/satijalab/seurat/releases) and [install from the tarball](https://stackoverflow.com/questions/4739837/how-do-i-install-an-r-package-from-the-source-tarball-on-windows). 

# Install the development version of Seurat

Install the development version of Seurat - directly from [GitHub](https://github.com/satijalab/seurat/tree/develop).

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
# Install the remotes package
install.packages('remotes')
remotes::install_github(repo = 'satijalab/seurat', ref = 'develop')
library(Seurat)
```

# Docker

We provide docker images for Seurat via [dockerhub](https://hub.docker.com/r/satijalab/seurat).

To pull the latest image from the command line:

```sh
docker pull satijalab/seurat:latest
```

To use as a base image in a new Dockerfile:

```sh
FROM satijalab/seurat:latest
```

