---
title: "Installation Instructions for Seurat"
output: html_document
---

To install Seurat, [R](https://www.r-project.org/) version 3.4 or greater is required. We also recommend installing [R Studio](https://www.rstudio.com/).

# Install from CRAN 
 
Seurat is available on [CRAN](https://cran.r-project.org/package=Seurat) for all platforms. To install, run: 

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
install.packages('Seurat')
library(Seurat)
```

If you see the warning message below, enter `y`:

```{r eval=FALSE}
package which is only available in source form, and may need compilation of C/C++/Fortran: 'Seurat'
Do you want to attempt to install these from sources?
y/n:
```

The CRAN release of Seurat is currently version 3.1.5. If you have installed version 3.1.5 and want to switch back to the last version 2, please see the "Previous Versions" section below 

# Install previous versions of Seurat

## Install the last version 2 release (2.3.4) 

To facilitate easy reinstallation of the last version 2 release, we are hosting the binaries on our website. These can be installed with the following command: 

```{r eval = FALSE}
source("https://z.umn.edu/archived-seurat")
```

 <details>
  <summary>View the script</summary>
  <script src="https://gist.github.com/satijalab/beb9bb50dedc75ee023bd5d9be5fe684.js"></script>
</details>

## Older versions of Seurat 
Old versions of Seurat, from Seurat v2.0.1 and up, are hosted in CRAN's archive. To install an old version of Seurat, run: 

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
# Install the remotes package 
install.packages('remotes')
# Replace '2.3.0' with your desired version
remotes::install_version(package = 'Seurat', version = package_version('2.3.0'))
library(Seurat)
```

For versions of Seurat older than those not hosted on CRAN (versions 1.3.0 and 1.4.0), please download the packaged source code from our [releases page](https://github.com/satijalab/seurat/releases) and [install from the tarball](https://stackoverflow.com/questions/4739837/how-do-i-install-an-r-package-from-the-source-tarball-on-windows). 

# Install the development version of Seurat

Install the development version of Seurat - directly from [GitHub](https://github.com/satijalab/seurat/tree/develop).

```{r eval = FALSE}
# Enter commands in R (or R studio, if installed)
# Install the remotes package
install.packages('remotes')
remotes::install_github(repo = 'satijalab/seurat', ref = 'develop')
library(Seurat)
```
