---
title: "Mapping and annotating query datasets"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  fig.width = 8,
  time_it = TRUE,
  error = TRUE
)
```

# Introduction to single-cell reference mapping

In this vignette, we first build an integrated reference and then demonstrate how to leverage this reference to annotate new query datasets. Generating an integrated reference follows the same workflow described in more detail in the integration introduction [vignette](integration_introduction.html). Once generated, this reference can be used to analyze additional query datasets through tasks like cell type label transfer and projecting query cells onto reference UMAPs. Notably, this does not require correction of the underlying raw query data and can therefore be an efficient strategy if a high quality reference is available. 

# Dataset preprocessing 

For the purposes of this example, we've chosen human pancreatic islet cell datasets produced across four technologies, CelSeq (GSE81076) CelSeq2 (GSE85241), Fluidigm C1 (GSE86469), and SMART-Seq2 (E-MTAB-5061). For convenience, we distribute this dataset through our [SeuratData](https://github.com/satijalab/seurat-data) package. The metadata contains the technology (`tech` column) and cell type annotations (`celltype` column) for each cell in the four datasets.

```{r libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratData)
library(ggplot2)
```

```{r install.data, eval=FALSE}
InstallData('panc8')
```

As a demonstration, we will use a subset of technologies to construct a reference. We will then map the remaining datasets onto this reference. We start by selecting cells from four technologies, and performing an analysis without integration.

```{r preprocessing1}
panc8 <- LoadData('panc8')
table(panc8$tech)

# we will use data from 2 technologies for the reference
pancreas.ref <- subset(panc8, tech %in% c("celseq2", "smartseq2"))
pancreas.ref[["RNA"]] <- split(pancreas.ref[["RNA"]], f = pancreas.ref$tech)

# pre-process dataset (without integration)
pancreas.ref <- NormalizeData(pancreas.ref)
pancreas.ref <- FindVariableFeatures(pancreas.ref)
pancreas.ref <- ScaleData(pancreas.ref)
pancreas.ref <- RunPCA(pancreas.ref)
pancreas.ref <- FindNeighbors(pancreas.ref, dims=1:30)
pancreas.ref <- FindClusters(pancreas.ref)
pancreas.ref <- RunUMAP(pancreas.ref, dims = 1:30)
DimPlot(pancreas.ref,group.by = c("celltype","tech"))
```

Next, we integrate the datasets into a shared reference. Please see our [introduction to integration vignette](https://satijalab.org/seurat/articles/integration_introduction)
```{r preprocessing3, fig.height = 4}
pancreas.ref <- IntegrateLayers(
  object = pancreas.ref, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = 'integrated.cca',
  verbose = FALSE)
pancreas.ref <- FindNeighbors(pancreas.ref,reduction='integrated.cca',dims=1:30)
pancreas.ref <- FindClusters(pancreas.ref)
pancreas.ref <- RunUMAP(pancreas.ref,reduction='integrated.cca',dims=1:30)
DimPlot(pancreas.ref,group.by = c("tech","celltype"))
```

# Cell type classification using an integrated reference

Seurat also supports the projection of reference data (or meta data) onto a query object. While many of the methods are conserved (both procedures begin by identifying anchors), there are two important distinctions between data transfer and integration:

1. In data transfer, Seurat does not correct or modify the query expression data.
2. In data transfer, Seurat has an option (set by default) to project the PCA structure of a reference onto the query, instead of learning a joint structure with CCA. We generally suggest using this option when projecting data between scRNA-seq datasets.

After finding anchors, we use the `TransferData()` function to classify the query cells based on reference data (a vector of reference cell type labels). `TransferData()` returns a matrix with predicted IDs and prediction scores, which we can add to the query metadata.

```{r label.transfer, warning = FALSE, message = FALSE}

# select two technologies for the query datasets
pancreas.query <- subset(panc8, tech %in% c("fluidigmc1","celseq"))
pancreas.query <- NormalizeData(pancreas.query)
pancreas.anchors <- FindTransferAnchors(reference = pancreas.ref, query = pancreas.query, dims = 1:30, reference.reduction="pca")
predictions <- TransferData(anchorset = pancreas.anchors, refdata = pancreas.ref$celltype, dims = 1:30)
pancreas.query <- AddMetaData(pancreas.query, metadata = predictions)
```

Because we have the original label annotations from our full integrated analysis, we can evaluate how well our predicted cell type annotations match the full reference. In this example, we find that there is a high agreement in cell type classification, with over 96% of cells being labeled correctly.

```{r analysis2}
pancreas.query$prediction.match <- pancreas.query$predicted.id == pancreas.query$celltype
table(pancreas.query$prediction.match)
```

To verify this further, we can examine some canonical cell type markers for specific pancreatic islet cell populations. Note that even though some of these cell types are only represented by one or two cells (e.g. epsilon cells), we are still able to classify them correctly. 

```{r vlnplots, fig.height=8}
table(pancreas.query$predicted.id)
VlnPlot(pancreas.query, c("REG1A", "PPY", "SST", "GHRL", "VWF", "SOX10"), group.by = "predicted.id") 
```

# Unimodal UMAP Projection

We also enable projection of a query onto the reference UMAP structure. This can be achieved by computing the reference UMAP model and then calling  `MapQuery()` instead of `TransferData()`. 

```{r label.transfer.v4, warning = FALSE, message = FALSE}
pancreas.ref <- RunUMAP(pancreas.ref, dims = 1:30, reduction = "integrated.cca", return.model = TRUE)
pancreas.query <- MapQuery(
  anchorset = pancreas.anchors, 
  reference = pancreas.ref,
  query = pancreas.query,
  refdata = list(celltype = 'celltype'),
  reference.reduction = "pca", 
  reduction.model = 'umap'
)
```

<details>
  <summary>**What is `MapQuery` doing?**</summary>

`MapQuery()` is a wrapper around three functions: `TransferData()`, `IntegrateEmbeddings()`, and `ProjectUMAP()`. `TransferData()` is used to transfer cell type labels and impute the ADT values; `IntegrateEmbeddings()` is used to integrate reference with query by correcting the query's projected low-dimensional embeddings; and finally `ProjectUMAP()` is used to project the query data onto the UMAP structure of the reference. The equivalent code for doing this with the intermediate functions is below:

```{r, eval=FALSE}
pancreas.query <- TransferData(
  anchorset = pancreas.anchors, 
  reference = pancreas.ref,
  query = pancreas.query,
  refdata = list(celltype = "celltype")
)
pancreas.query <- IntegrateEmbeddings(
  anchorset = pancreas.anchors,
  reference = pancreas.ref,
  query = pancreas.query, 
  new.reduction.name = "ref.pca"
)
pancreas.query <- ProjectUMAP(
  query = pancreas.query, 
  query.reduction = "ref.pca", 
  reference = pancreas.ref, 
  reference.reduction = "pca", 
  reduction.model = "umap"
)
```
</details>

We can now visualize the query cells alongside our reference.

```{r panc.refdimplots, fig.width=10}
p1 <- DimPlot(pancreas.ref, reduction = "umap", group.by = "celltype", label = TRUE,
             label.size = 3 ,repel = TRUE) + NoLegend() + ggtitle("Reference annotations")
p2 <- DimPlot(pancreas.query, reduction = "ref.umap", group.by = "predicted.celltype", label = TRUE, 
             label.size = 3 ,repel = TRUE) + NoLegend() + ggtitle("Query transferred labels")
p1 + p2
```

```{r save.times, include = FALSE}
write.csv(x = t(as.data.frame(all_times)), file = "../output/timings/integration_reference_mapping.csv")
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
