---
title: "Documentation Archive"
output:
  html_document:
    theme: united
    df_print: kable
---

```{r helper, include = FALSE}
make_vignette_button <- function(name, title, version) {
  url <- paste0("../archive/", version, "/", name, ".html")
  paste0('<a class="btn btn-primary" style="margin-right:1rem" href=', url ,'>', version, '</a>')
}
```

```{r yaml, include = FALSE}
library(yaml)
vdat <- read_yaml(file = "archive.yaml")
```
In version 4, the Seurat documentation was transitioned to pkgdown. Here we provide access to all previous versions of the documentation.

# Version 2-3 tutorials

```{r results='asis', echo=FALSE, warning=FALSE, message = FALSE}
for (i in 1:length(x = vdat$vignettes)) {
  vignette <- vdat$vignettes[[i]]
  cat('<h3>', vignette$title, '</h3>')
  for(j in 1:length(x = vignette$versions)) {
    cat(make_vignette_button(name = vignette$name, title = vignette$title, version = vignette$version[j]))
  }
}
```

# Version 1.3-1.4 tutorials

For versions 1.3-1.4, we provide access to the old documentation pages [here](../archive/v1.4/get_started_v1_4.html)

# Version <=1.2 tutorials

For versions <=1.2, we provide access to the old documentation pages [here](../archive/v1.2/get_started_v1_2.html)

