---
title: "Seurat v5 Command Cheat Sheet"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---
***
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = 'styler',
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  fig.width = 8,
  time_it = TRUE
)
```

Here, we describe important commands and functions to store, access, and process data using Seurat v5. To demonstrate commamnds, we use a dataset of 3,000 PBMC (stored in-memory), and a dataset of 1.3M E18 mouse neurons (stored on-disk), which we constructed as described in the [BPCells vignette](seurat5_bpcells_interaction_vignette.html).

```{r loaddata}
library(Seurat)
library(SeuratData)
library(BPCells)
library(dplyr)
options(Seurat.object.assay.version = "v5")
```

## Load datasets

```{r matrix}
pbmc3k <- LoadData("pbmc3k")
mousebrain1m <- readRDS("/brahms/hartmana/vignette_data/1p3_million_mouse_brain.rds")

#  RNA assay is of the Assay5 class
class(pbmc3k[["RNA"]])
class(mousebrain1m[["RNA"]])
```

## Access and store expression data

The `$` and double-bracket `[[]]` symbols can be used as efficient accessor functions for Seurat5 assays.

```{r }

# access the counts matrix from the RNA assay
counts_matrix <- pbmc3k[["RNA"]]$counts

# Add a layer
# Equivalent to running pbmc3k <-NormalizeData(pbmc3k)
pbmc3k[["RNA"]]$data <- NormalizeData(pbmc3k[["RNA"]]$counts)

# Delete a layer
pbmc3k[["RNA"]]$data <- NULL

# pbmc3k counts matrix is stored in-memory
class(pbmc3k[["RNA"]]$counts)

# 1.3M cell dataset counts matrix is stored on-disk
class(mousebrain1m[["RNA"]]$counts)
```

Despite the drastic difference in dataset size, the 1.3M cell dataset occupies a small memory footprint thanks to on-disk storage.

```{r}
paste("PBMC 3k contains", length(colnames(pbmc3k)), "cells")
paste("Mouse brain 1.3M contains", length(colnames(mousebrain1m)), "cells")

# Despite the mouse brain dataset containing 1.3 million cells, the assay is under 350Mbs in size due to on-disk storage
paste("PBMC 3k assay size:", format(object.size(pbmc3k[["RNA"]]), units = "Mb"))
paste("Mouse brain 1.3M assay size:", format(object.size(mousebrain1m[["RNA"]]), units = "Mb"))
```

## Access cell names and metadata

Get cell names. Since Seurat v5 object doesn't require all assays have the same cells, `Cells()` is designed to get cell names of the default assay and `colnames()` is deigned to get cell names of the entire object

```{r}
pbmc3k[["RNAsub"]] <- subset(pbmc3k[["RNA"]], cells = colnames(pbmc3k)[1:100])
DefaultAssay(pbmc3k) <- 'RNAsub'
length(Cells(pbmc3k))
length(colnames(pbmc3k))
```

Access object metadata

```{r meta}

# get all object metadata
pbmc_metadata <- pbmc3k[[]]

# get list of metadata columns
colnames(pbmc_metadata)

# get annotations stored in metadata
annotations <- pbmc3k$seurat_annotations

```

## Create Seurat or Assay objects

By setting a global option (`Seurat.object.assay.version`), you can default to creating either Seurat v3 assays, or Seurat v5 assays. The use of v3 assays is set by default upon package loading, which ensures backwards compatibiltiy with existing workflows.

```{r create}
# create v3 assays
options(Seurat.object.assay.version = "v3")
pbmc.counts <- Read10X(data.dir = "/brahms/hartmana/vignette_data/pbmc3k/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.counts)
class(pbmc[["RNA"]])

# create v5 assays
options(Seurat.object.assay.version = "v5")
pbmc.counts <- Read10X(data.dir = "/brahms/hartmana/vignette_data/pbmc3k/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.counts)
class(pbmc[["RNA"]])
```

`CreateAssayObject()` and `CreateAssay5Object()` can be used to create v3 and v5 assay regardless of the setting in `Seurat.object.assay.version`

```{r}
#create a v3 assay
assay.v3 <- CreateAssayObject(counts = pbmc.counts)

#create a v5 assay
assay.v5 <- CreateAssay5Object(counts = pbmc.counts)

class(assay.v3)
class(assay.v5)
```

Assay5 objects are more flexible, and can be used to store only a data layer, with no counts data. This can be used to create Seurat objects that require less space

```{r}
# create an assay using only normalized data
assay.v5 <- CreateAssay5Object(data = log1p(pbmc.counts))

# create a Seurat object based on this assay
pbmc3k_slim <- CreateSeuratObject(assay.v5)
pbmc3k_slim
```

We can also convert (cast) between `Assay` and `Assay5` objects with `as()`. 

```{r}
# convert a v5 assay to a v3 assay
pbmc3k[["RNA3"]] <- as(object = pbmc3k[["RNA"]], Class = "Assay")

# convert a v3 assay to a v5 assay
pbmc3k[["RNA5"]] <- as(object = pbmc3k[["RNA3"]], Class = "Assay5")
```

## Working with layers

Seurat v5 assays store data in layers. These layers can store raw, un-normalized counts (`layer='counts'`), normalized data (`layer='data'`), or z-scored/variance-stabilized data (`layer='scale.data'`). 

```{r}
# by default, creates an RNA assay with a counts layer
obj <- CreateSeuratObject(counts = pbmc.counts)
obj

# creates a normalized data layer
obj <- NormalizeData(obj,verbose = FALSE)
obj

#extract only the layer names from an assay
Layers(obj[["RNA"]])
```

Prior to performing integration analysis in Seurat v5, we can split the layers into groups. The `IntegrateLayers` function, described in [our vignette](seurat5_integration.html), will then align shared cell types across these layers. After performing integration, you can rejoin the layers.

```{r joinsplit}
# create random batches
pbmc3k$batch <- sample(c("batchA","batchB","batchC"),ncol(pbmc3k),replace = TRUE)

# split layers
pbmc3k[["RNA"]] <- split(pbmc3k[["RNA"]], f=pbmc3k$batch)
Layers(pbmc3k[["RNA"]])

# rejoin layers
pbmc3k[["RNA"]] <- JoinLayers(pbmc3k[["RNA"]])
Layers(pbmc3k[["RNA"]])
```

If you have multiple counts matrices, you can also create a Seurat object that is initialized with multiple layers.

```{r multilayer}
batchA_counts <- pbmc.counts[,1:200]
batchB_counts <- pbmc.counts[,201:400]
batchC_counts <- pbmc.counts[,401:600]
count_list <- list(batchA_counts,batchB_counts,batchC_counts)
names(count_list) <- c('batchA','batchB','batchC')

# create a Seurat object initialized with multiple layers
obj <- CreateSeuratObject(counts = count_list)
Layers(obj[["RNA"]])
```

## Accessing additional data

```{r}
pbmc3k <- FindVariableFeatures(pbmc3k, verbose = FALSE)
pbmc3k <- ScaleData(pbmc3k,verbose = FALSE)
pbmc3k <- RunPCA(pbmc3k,verbose = FALSE)

# return variable features

# returns information from both assay, cell embeddings and meta.data as a data.frame
fetch_df <- FetchData(object = pbmc3k, layer = "counts", vars = c("rna_MS4A1", "PC_1", "nCount_RNA"))
head(fetch_df)

# get cell embeddings
head(Embeddings(object = pbmc3k[['pca']])[, 1:5])

# get feature loadings
head(Loadings(object  = pbmc3k[['pca']])[, 1:5])
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
